---
title: "Precipitation Interpolation"
author: "Matthew Ross"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(sf) # new goodness
library(mapview)
library(lubridate)
library(osmdata)
library(gstat)
library(elevatr)
library(riem)

knitr::opts_chunk$set(echo = TRUE)
```



# Read in Daily P data for all CO ASOS sites

See previous assignments or videos for description of ASOS data

```{r}

load('data/DailyP.RData')

head(daily_p)
```


## Get site elevation

```{r, eval = F}
unique_asos <- daily_p %>%
  distinct(station, lon, lat) %>%
  st_as_sf(., coords = c('lon','lat'),crs = 4326) %>%
  get_elev_point(.)


st_write(unique_asos,'data/unique_asos_elev.gpkg')
```


## Take station month averages


```{r}

unique_asos <- st_read('data/unique_asos_elev.gpkg')

annual_p <- daily_p %>%
  mutate(month = month(date)) %>%
  group_by(month,station,lon,lat) %>%
  summarize(across(where(is.numeric), sum)) %>%
  left_join(unique_asos)



```



## Look at  (Precip ~ elevation) for All months




```{r}

ggplot(annual_p,aes(x=elevation, y=daily_p,color = month)) + 
  geom_point() + 
  ylab('Monthly Precipitation (in)') + 
  scale_color_viridis_c()



```



## Pick a month (summer months are safer)

### Build IDW precip for state for that month



### Build IDW with elevation for state for that month including elevation as a 
variable

```{r}
## You will need to create a Stars raster that has elevation data. 



```


### Compare both maps to PRISM approach for your month


How close do our simple approaches come to reproducing prism maps? 


https://www.prism.oregonstate.edu/recent/monthly.php
