---
title: "Precipitation Interpolation"
author: "Matthew Ross"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(sf) # new goodness
library(mapview)
library(lubridate)
library(osmdata)
library(gstat)
library(elevatr)


knitr::opts_chunk$set(echo = TRUE)
```


# Load in Precip data

```{r}
load('data/DailyP.RData')

head(daily_p)
```


## Get Elevation Data

```{r, eval = F}
unique_asos <- daily_p %>%
  distinct(lon, lat, station)  %>%
  st_as_sf(., coords = c('lon','lat'), crs = 4326) %>%
  get_elev_point(.)

st_write(unique_asos, 'data/unique_asso_elev.gpkg')
```

## Get Monthly P Averages


```{r}
monthly_p <- daily_p %>%
  mutate(month = month(date)) %>%
  group_by(month, station) %>%
  summarize(monthly_p = sum(daily_p)) %>%
  left_join(unique_asos) #grab elevation data
  

```


## Look at monthly P

```{r}
ggplot(monthly_p, aes(x = elevation, y = monthly_p, color = month)) + 
  scale_color_viridis_c() + 
  geom_point()
```


## Getting Monthly Means of means, mins, maxes. 

```{r}
monthly_t <- daily_p %>%
  mutate(month = month(date)) %>%
  group_by(month, station) %>%
  select(-lon,-lat) %>%
  summarize(across(where(is.numeric), mean)) %>%
  left_join(unique_asos)


```


## Temp vs Elevation


```{r}
ggplot(monthly_t, aes(y = mean_temp, x = elevation, color = month)) + 
  geom_point() + 
  scale_color_viridis_c()
```



## Pick a month (summer months are safer)


```{r}

```

### Build IDW precip or elevation for state for that month

```{r}

```

### Plot this data

```{r}

```

### Build IDW with elevation for state for that month including elevation as a predictor

Hint! Use `get_elev_raster` 

```{r}

```

### Make a Map of that

```{r}
## You will need to create a Stars raster that has elevation data. 



```


### Compare both maps to PRISM approach for your month


How close do our simple approaches come to reproducing prism maps? 


https://www.prism.oregonstate.edu/recent/monthly.php




```{r}

```

