---
title: "Precipitation Interpolation"
author: "Matthew Ross"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(sf) # new goodness
library(mapview)
library(lubridate)
library(osmdata)
library(gstat)
library(elevatr)

knitr::opts_chunk$set(echo = TRUE)
```


# Load in Precip data

```{r}

load("data/DailyP.Rdata")


```

# Get elevation data

```{r}
#pulling the unqiue metadata info
unique_asos <- daily_p %>%
  distinct(lon, lat, station) %>%
  select(lon, lat, station) %>% #reordering to match get_elev_point
  st_as_sf(., coords = c('lon','lat'), crs = 4326) %>% 
  get_elev_point(.)

st_write(unique_asos, 'data/unique_asos_elev.gpkg')


```

# Get monthly P Averages

```{r}

monthly_p <- daily_p %>%
  mutate(month = month(date)) %>%
  group_by(month, station) %>%
  summarize(monthly_p = sum(daily_p)) %>%
  left_join(unique_asos) #grab elevation data
  

```

#look at monthly P

```{r}
ggplot(monthly_p, aes(x = elevation, y = monthly_p, color = month)) +
  geom_point() +
  scale_color_viridis_c()

```
#get monthly mean temp of means, mins and maxes

```{r}
monthly_t <- daily_p %>%
  mutate(month = month(date)) %>%
  group_by(month, station) %>%
  select(-c(lon,lat)) %>% 
  summarize(across(where(is.numeric), mean)) %>%
  left_join(unique_asos, .)



```
# temp vs. elev

```{r}
ggplot(monthly_t, aes(x = elevation, y = mean_temp, color = month)) +
  geom_point() +
  scale_color_viridis_c()
```

## Pick a month (summer months are safer)


```{r}
#going to pick a summer month for temp
unique_asos <- unique_asos %>%
  st_transform(2163)

aug_t <- monthly_t %>%
  filter(month == 8) %>%
  na.omit() %>% 
  st_transform(2163)
  
```

### Build IDW precip or elevation for state for that month

```{r}
library(stars)

#bb <- get_elev_raster(unique_asos, z=7)

bb <- st_bbox(unique_asos) %>%
  st_as_stars(dx = 10000)

mapview(bb)

interp = idw(elevation ~ 1, aug_t, bb)

#mapview (interp) + mapview(aug_t, zcol='elevation')

```

### Plot this data

```{r}
mapview (interp) + mapview(aug_t, zcol='elevation')

```

### Build IDW with elevation for state for that month including elevation as a predictor

Hint! Use `get_elev_raster` 

```{r}



```

### Make a Map of that

```{r}
## You will need to create a Stars raster that has elevation data. 
```


### Compare both maps to PRISM approach for your month


How close do our simple approaches come to reproducing prism maps? 


https://www.prism.oregonstate.edu/recent/monthly.php




```{r}
```