---
title: "Precipitation Interpolation"
author: "Matthew Ross"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(sf) # new goodness
library(mapview)
library(lubridate)
library(osmdata)
library(gstat)
library(elevatr)

knitr::opts_chunk$set(echo = TRUE)
```


# Load in Precip data

```{r}

load("data/DailyP.Rdata")

```

# Get elevation data

```{r}
#pulling the unqiue metadata info
unique_asos <- daily_p %>%
  distinct(lon, lat, station) %>%
  select(lon, lat, station) %>% #reordering to match get_elev_point
  st_as_sf(., coords = c('lon','lat'), crs = 4326) %>% 
  get_elev_point(.)

#st_write(unique_asos, 'data/unique_asos_elev.gpkg')


```

# Get monthly P Averages

```{r}

monthly_p <- daily_p %>%
  mutate(month = month(date)) %>%
  group_by(month, station) %>%
  summarize(monthly_p = sum(daily_p)) %>%
  left_join(unique_asos) #grab elevation data
  

```

#look at monthly P

```{r}
ggplot(monthly_p, aes(x = elevation, y = monthly_p, color = month)) +
  geom_point() +
  scale_color_viridis_c()

```
#get monthly mean temp of means, mins and maxes

```{r}
monthly_t <- daily_p %>%
  mutate(month = month(date)) %>%
  group_by(month, station) %>%
  select(-c(lon,lat)) %>% 
  summarize(across(where(is.numeric), mean)) %>%
  left_join(unique_asos, .)

```
# temp vs. elev

```{r}
ggplot(monthly_t, aes(x = elevation, y = mean_temp, color = month)) +
  geom_point() +
  scale_color_viridis_c()
```

## Pick a month (summer months are safer)


```{r}
#going to pick a summer month for temp
unique_asos <- unique_asos %>%
  st_transform(2163)

aug_t <- monthly_t %>%
  filter(month == 8) %>%
  na.omit() %>% 
  st_transform(2163)
  
```

### Build IDW precip or elevation for state for that month

```{r}
library(stars)
library(tmap)

#bb <- get_elev_raster(unique_asos, z=7)

bb <- st_bbox(unique_asos) %>%
  st_as_stars(dx = 1000)

mapview(bb) #do plot(bb) bc it is projecting differently on mapview

interp = idw(daily_p ~ 1, aug_t, bb)

mapview(interp)

plot(interp)

tm_shape(interp[1]) +
  tm_raster(palette = 'Blues',
            style='cont')

```

### Plot this data

```{r}
mapview (interp) + mapview(aug_t, zcol='elevation')

```

### Build IDW with elevation for state for that month including elevation as a predictor

Hint! Use `get_elev_raster` 

```{r}

elev <- get_elev_raster(aug_t, z=6)

elev_stars <- st_as_stars(elev, dx=1000) #%>%
  #st_transform(2163)

mapview(elev_stars)

names(elev_stars) <- 'elevation'

interp2 = idw(daily_p ~ elevation, aug_t, elev_stars)


```

### Make a Map of that

```{r}
## You will need to create a Stars raster that has elevation data. 

mapview(interp2)

tm_shape(interp2[1]) +
  tm_raster(palette = "Blues",
            style = 'cont')


```

### Compare both maps to PRISM approach for your month


How close do our simple approaches come to reproducing prism maps? 


https://www.prism.oregonstate.edu/recent/monthly.php

```{r}
```

## ANSWER KEY BELOW


### Build IDW precip or elevation for Colorado for that month

```{r}
unique_asos_2163 <- st_transform(unique_asos,crs = 2163)
co_box <- st_bbox(unique_asos_2163) %>%
  st_as_stars(dx = 1000)
july_t <- monthly_t %>%
  dplyr::filter(month == 6) %>%
  st_transform(., st_crs(co_box)) %>%
  na.omit(.)
interp_basic = idw(mean_temp~1, july_t, co_box) %>%
  dplyr::select(1)
tm_shape(interp_basic) + 
  tm_raster(palette = 'Reds', style = 'cont')
```

### Plot this data

```{r}
```

### Build IDW with elevation for state for that month including elevation as a predictor

Hint! Use `get_elev_raster` 

```{r}
library(raster)
ras <- get_elev_raster(unique_asos, z = 7) %>%
  raster::crop(.,unique_asos)
co_stars <- st_as_stars(ras)
names(co_stars) <- 'elevation'
june_t <- monthly_t %>%
  filter(month == 6)
ggplot(june_t, aes(elevation, mean_temp)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
interp = gstat::idw(mean_temp~elevation, 
                    june_t, 
                    co_stars) %>%
  dplyr::select(1)
```

### Make a Map of that

```{r}
## You will need to create a Stars raster that has elevation data. 
tm_shape(interp) + 
  tm_raster(palette = 'plasma', style = 'cont')
mapview(interp)